A quantitative framework for investigating the reliability of empirical network construction: supplemental information

### 2019-03-13 manuscript/CI_interactions.tex
commented out figures for manuscriptcentral
resubmitted second round March 5
working through SI

All data used in this study have been independently published and are accessible following the references provided in text.

@@

  \begin{figure}[h!]
    \caption{Nested levels of uncertainty mean that an observed interaction matrix is unlikely to capture all of the interactions that truly occur. Some feasible interactions (black squares in True matrix) are less likely than others (grey squares; lighter grey is less likely) based on species traits. For example, lions are less likely to predate upon elephants than zebras. If our trait model is incomplete (e.g., if we neglect group hunting in lions), we might assume that elephants are too large to be lion prey and assign this interaction a probability of 0.\\
    \indent Interactions occur with different probabilities in any given sample. We show six example networks representing interactions observed over two days at three sites. Interactions which occur during fewer sampling days are less likely to be observed.\\
    \indent Not all interactions are equally detectable. For example, interactions involving cryptic species are less likely to be observed.\\
    \indent After combining these layers of uncertainty, some interactions which truly occur are very unlikely to be observed. The observed matrix (bottom) is a subsample of the true matrix where the probability of observing each interaction depends upon multiple layers of uncertainty. Note that some low-probability interactions (e.g., lion predation on zebra) are included while others (e.g., lion predation on elephants) are not. For an illustration of the way in which these levels of uncertainty combine in simulated data, and the effects of uncertainty on network structure, see \emph{Appendix S1}. Attributions for images are given in \emph{Appendix S1-A}.}
    \label{conceptual_fig}
    \begin{center}
    \includegraphics*[height=.45\textheight]{figures/conceptual_fig.eps}
    % \includegraphics*[height=.5\textheight]{Figure_1.eps}
    \end{center}
    \end{figure}

  \begin{figure}[h!]
      \caption{Despite the high replication in our empirical dataset, most galler-natural enemy pairs were never observed co-occurring and those that co-occurred rarely interacted. \textbf{A)} Histogram of the number of pairs of species observed co-occurring at least once. 9794 galler-enemy pairs were never observed co-occurring (not shown).
      \textbf{B)} Histogram of the number of observed interactions between co-occurring species. Species which co-occurred but never interacted are included. 
      \textbf{C)} For each species pair, the number of observed interactions plotted against the number of observed co-occurrences. The red, dashed line indicates a 1:1 relationship.}
      \label{histograms}
      \begin{center}
      \includegraphics*[height=.6\textheight]{figures/GP_histogram.eps}
      % \includegraphics*[height=.6\textheight]{Figure_2.eps}
      \end{center}
      \end{figure}

  \begin{figure}[h!]
    \caption{Posterior distributions for the interaction probability ($\lambda_{ij}$) for two species that have not been observed interacting ($k_{ij}=0$), based on a single site (Zillis in Graub\"{u}nden, Switzerland) from~\citet{Kopelke2017}. Posterior distributions (curves) and 95\% credible intervals (lines at top of panel) for $\lambda_{ij}$ narrow and approach zero as the number of observed co-occurrences ($n_{ij}$) increases. Diamonds indicate the maximum likelihood estimator for the mean probability of interaction.
    Dashed lines indicate threshold probabilities of 0.01, 0.05, and 0.1. The number of samples required to obtain a 95\% credible interval below each threshold increases rapidly. It takes just over 20 observed co-occurrences to be 95\% confident that $\lambda_{ij}<0.10$ but over 100 co-occurrences to be 95\% confident that $\lambda_{ij}<0.01$.}
    \label{Salix_pdfs_cdfs}
    \begin{center}
    \includegraphics*[width=.8\textwidth]{figures/GP_pdfs_increasing_N_Zillis.eps}
    % \includegraphics*[width=.8\textwidth]{Figure_3.eps}
    \end{center}
    \end{figure}

  \begin{figure}[h!]
    \caption{Posterior distributions for the interaction probability ($\lambda_{ij}$) for two species that have not been observed interacting ($k_{ij}=0$), A) based on a single site (Zillis in Graub\"{u}nden, Switzerland) from~\citet{Kopelke2017} and B) based on a network describing interactions between gallers and several genotypes of \emph{Salix hookeriana} from~\citet{Barbour2016}. Posterior distributions (curves) and 95\% credible intervals (lines at top of panel) for $\lambda_{ij}$ narrow and approach zero as the number of observed co-occurrences ($n_{ij}$) increases. Diamonds indicate the maximum likelihood estimator for the mean probability of interaction.
    Dashed lines indicate threshold probabilities of 0.01, 0.05, and 0.1. Note that both priors constrain the MLE of $\lambda_{ij}$ above 0.1 and that the posterior probability distributions become similar for very high $n$.}
    \label{prior_comparison}
    \begin{center}
    \includegraphics*[width=.8\textwidth]{figures/SG_pdfs_prior_comparison.eps}
    % \includegraphics*[width=.8\textwidth]{Figure_4.eps}
    \end{center}
    \end{figure}

>>>>

Data from~\citet{Kopelke2017} used in this manuscript is available from \emph{https://doi.org/10.1002/ecy.1832}. Data from~\citet{Barbour2016} used in this manuscript is available from \emph{https://doi.org/10.5061/dryad.g7805}. All code used in this manuscript is available in the online supporting information.

@@

  \begin{figure}[h!]
    \caption{Nested levels of uncertainty mean that an observed interaction matrix is unlikely to capture all of the interactions that truly occur. Some feasible interactions (black squares in True matrix) are less likely than others (grey squares; lighter grey is less likely) based on species traits. For example, lions are less likely to predate upon elephants than zebras. If our trait model is incomplete (e.g., if we neglect group hunting in lions), we might assume that elephants are too large to be lion prey and assign this interaction a probability of 0.\\
    \indent Interactions occur with different probabilities in any given sample. We show six example networks representing interactions observed over two days at three sites. Interactions which occur during fewer sampling days are less likely to be observed.\\
    \indent Not all interactions are equally detectable. For example, interactions involving cryptic species are less likely to be observed.\\
    \indent After combining these layers of uncertainty, some interactions which truly occur are very unlikely to be observed. The observed matrix (bottom) is a subsample of the true matrix where the probability of observing each interaction depends upon multiple layers of uncertainty. Note that some low-probability interactions (e.g., lion predation on zebra) are included while others (e.g., lion predation on elephants) are not. For an illustration of the way in which these levels of uncertainty combine in simulated data, and the effects of uncertainty on network structure, see \emph{Appendix S1}. Attributions for images are given in \emph{Appendix S1-A}.}
    \label{conceptual_fig}
    \begin{center}
    % \includegraphics*[height=.45\textheight]{figures/conceptual_fig.eps}
    % \includegraphics*[height=.5\textheight]{Figure_1.eps}
    \end{center}
    \end{figure}

  \begin{figure}[h!]
      \caption{Despite the high replication in our empirical dataset, most galler-natural enemy pairs were never observed co-occurring and those that co-occurred rarely interacted. \textbf{A)} Histogram of the number of pairs of species observed co-occurring at least once. 9794 galler-enemy pairs were never observed co-occurring (not shown).
      \textbf{B)} Histogram of the number of observed interactions between co-occurring species. Species which co-occurred but never interacted are included. 
      \textbf{C)} For each species pair, the number of observed interactions plotted against the number of observed co-occurrences. The red, dashed line indicates a 1:1 relationship.}
      \label{histograms}
      \begin{center}
      % \includegraphics*[height=.6\textheight]{figures/GP_histogram.eps}
      % \includegraphics*[height=.6\textheight]{Figure_2.eps}
      \end{center}
      \end{figure}

  \begin{figure}[h!]
    \caption{Posterior distributions for the interaction probability ($\lambda_{ij}$) for two species that have not been observed interacting ($k_{ij}=0$), based on a single site (Zillis in Graub\"{u}nden, Switzerland) from~\citet{Kopelke2017}. Posterior distributions (curves) and 95\% credible intervals (lines at top of panel) for $\lambda_{ij}$ narrow and approach zero as the number of observed co-occurrences ($n_{ij}$) increases. Diamonds indicate the maximum likelihood estimator for the mean probability of interaction.
    Dashed lines indicate threshold probabilities of 0.01, 0.05, and 0.1. The number of samples required to obtain a 95\% credible interval below each threshold increases rapidly. It takes just over 20 observed co-occurrences to be 95\% confident that $\lambda_{ij}<0.10$ but over 100 co-occurrences to be 95\% confident that $\lambda_{ij}<0.01$.}
    \label{Salix_pdfs_cdfs}
    \begin{center}
    % \includegraphics*[width=.8\textwidth]{figures/GP_pdfs_increasing_N_Zillis.eps}
    % \includegraphics*[width=.8\textwidth]{Figure_3.eps}
    \end{center}
    \end{figure}

  \begin{figure}[h!]
    \caption{Posterior distributions for the interaction probability ($\lambda_{ij}$) for two species that have not been observed interacting ($k_{ij}=0$), A) based on a single site (Zillis in Graub\"{u}nden, Switzerland) from~\citet{Kopelke2017} and B) based on a network describing interactions between gallers and several genotypes of \emph{Salix hookeriana} from~\citet{Barbour2016}. Posterior distributions (curves) and 95\% credible intervals (lines at top of panel) for $\lambda_{ij}$ narrow and approach zero as the number of observed co-occurrences ($n_{ij}$) increases. Diamonds indicate the maximum likelihood estimator for the mean probability of interaction.
    Dashed lines indicate threshold probabilities of 0.01, 0.05, and 0.1. Note that both priors constrain the MLE of $\lambda_{ij}$ above 0.1 and that the posterior probability distributions become similar for very high $n$.}
    \label{prior_comparison}
    \begin{center}
    % \includegraphics*[width=.8\textwidth]{figures/SG_pdfs_prior_comparison.eps}
    % \includegraphics*[width=.8\textwidth]{Figure_4.eps}
    \end{center}
    \end{figure}

### 2019-03-13 manuscript/supplemental.tex
working through SI

\author{Alyssa R. Cirtwill$^{1\dagger}$, \&  Anna Ekl\"{o}f$^{1}$, Tomas Roslin$^{2}$, Kate Wootton$^{2}$, Dominique Gravel$^{3}$}
\date{\small$^1$Department of Physics, Chemistry\\ 
and Biology (IFM)\\ 
Link\"{o}ping University\\
Link\"{o}ping, Sweden\\
\medskip $^2$ Department of Ecology\\ P.O. Box 7044, Swedish University of Agricultural Sciences \\ SE-750 07 Uppsala, Sweden \\
\medskip$^3$ D\'{e}partement de biologie\\ 
Universit\'{e} de Sherbrooke\\ 
Sherbrooke, Canada\\ 
\medskip
$^\dagger$ Corresponding author:\\
alyssa.cirtwill@gmail.com\\
+46 723 158464\\
}
\renewcommand\Authands{ and }

@@

\linenumbers

@@

  To demonstrate the effects of nested interaction, process, and detection uncertainty on network structure (see Fig. 1, main text), we can observe how the structure of randomly-generated networks changes as we add layers of uncertainty. We began by constructing 100 ``true" networks with 50 consumer species, 50 resource species, and a connectance of 0.1. We ensured that each network had at least one link per consumer and per resource. For each network, we calculated the mean number of links per consumer, mean number of links per resource, and nestedness (NODF) as described in the main text. For simplicity, we assume that there is no interaction uncertainty in this system (i.e., all interactions are either 100\% feasible or 100\% unfeasible based on species' traits).

@@

  With each additional layer of uncertainty, the number of consumers and resources included in the web decreased (mean 49.8 of each species type in detection-filter webs and mean 48.6 of each type in process-filtered webs). The connectance and number of links per consumer and per resource decreased significantly at each level (Fig.~\ref{sim_props}A,B,D; $p$\textless0.001 in all cases based on one-tailed Student's t tests). The effects of uncertainty on nestedness were less predictable, particularly in the detection-filtered webs where variability was much higher than in other web types. Nevertheless, mean nestedness was lower in detection-filtered webs than in the original webs ($t$=19.9, $p$\textless0.001) and lower in the process-filtered webs than the original and detection-filtered webs (Fig.~\ref{sim_props}C; $t$=32.8, $p$\textless0.001 and $t$=93.7, $p$\textless0.001). We note that these tests may reflect the large number of detection-filtered webs generated by our simulation. A single observed network drawn from these detection-filtered webs might well have a higher nestedness than the ``true" network. This simulation demonstrates the fact that process and detection uncertainty mean that we are very unlikely to observe all of the interactions that can truly occur in a system. It also highlights the sometimes unpredictable effects that this uncertainty can have on higher-order structures such as nestedness.

@@

        The obvious rule of thumb to reduce uncertainty is ``sample more". Where resources permit, increasing sample sizes will reduce uncertainty about the upper bound of the interaction probability and will also increase the probability of detecting unlikely or cryptic interactions (e.g., interactions where $L_{ij} = 1$ but process or detection uncertainty is high). Despite these benefits, we note that there are limits to the utility of increased sampling. Since the probability of observing the co-occurrence of two species will always be higher than the probability of observing their interaction (since the probability of interaction is conditional on both interaction partners being present~\citep{Gravel2018}), we will accumulate observations of co-occurrences faster than we will accumulate observations of interactions (Fig. 2C, main text). While this improves our understanding of the set of species present in the community, it introduces yet more uncertainty into the matrix of feasible interactions. Note that it is better to identify the set of potential interactions than to miss a species, so sampling effort should not be reduced in order to exclude rare species.

@@

        \vspace{12pt}
        \noindent\emph{
          degree=c(14, 4, 2, 3, 17, 6, 2, 15, 1, 1)
            }.
          \vspace{12pt}

@@

        \vspace{12pt}
        \noindent\emph{
           int\_probs = c(0.358, 0.110, 0.055, 0.082, 0.468, 0.165, 0.055, 0.412, 0.028, 0.028) }.
          \vspace{12pt}

        The mean of these interaction probabilities is 0.176, approximately two-thirds the $\lambda_{ij,MLE}$ obtained from the observed data. We can use the distribution of these interaction probabilities as our prior distribution and estimate the uncertainty surrounding our $\lambda_{ij,MLE}$. With some simple R code (function ``calculate\_parameters", \emph{Appendix S7}), we obtain prior parameters $\alpha$=0.998 and $\beta$=4.63. Using these priors in equations~\ref{mean} and~\ref{variance} above (or in the R function ``calculate\_distribution" in \emph{Appendix S7}), we find a prior $\bar\lambda_{ij}$=0.177 and var($\lambda_{ij}$)=0.026. Adding the observed data ($n_{ij}=10$, $k_{ij}=3$) and using the same code, we obtain posterior parameters $\alpha'$=4.00 and $\beta'$=11.6 and a posterior $\bar\lambda_{ij}$=0.256 and var($\lambda_{ij}$)=0.012. Comparing the posterior distribution to the prior, we see that the posterior is closer to the observed data and that the additional data about interactions between species $i$ and $j$ has reduced the variance. We may also wish to calculate a credible interval (analogous to the frequentist confidence interval). This is also quite straightforward in R (see function ``credible\_interval" in \emph{Appendix S7}). In this case, a 95\% credible interval for $\bar\lambda_{ij}$ is (0.080, 0.491).

@@

      The \emph{Salix}-galler-parasitoid meta-network dataset collected 
      by~\citet{Kopelke2017} consists of a single community type sampled across 
      Europe: willow (\emph{Salix}) species, willow-galling sawflies (Hymenoptera, Tenthredinidae, Nematinae, Euurina), and their natural 
      enemies (hymenopteran parasitoids and coleopteran, lepidopteran, dipteran, 
      and hymenopteran inquilines). The data were collected over 29 years 
      (1982-2010) at 374 unique locations across Europe ranging from Sicily to 
      the Arctic. Since the dataset includes repeated visits to a number of sites, the data set 
      contains 641 site-visits. Here we take the more conservative approach and pool visits to the same site for a sample size of 374 sub-networks.

      The meta-network consists of 1,173 different interactions 
      between 52 \emph{Salix} nodes, 92 herbivore nodes, and 126 parasitoid 
      nodes. Interactions were determined by dissecting and rearing gall 
      inhabitants from 165,424 galls. Some sites were visited repeatedly, for a 
      total of 641 site-visits. We consider these to be repeated samples and 
      take the 374 unique sites as our sample size.

@@

      For species where $n_{ij}=0$ (3,986/4,992 \emph{Salix}-galler pairs in our dataset), estimates for the mean and variance of $\lambda_{ij}$ can be calculated directly from the prior distribution. For the \emph{Salix}-galler network, these parameters were: $\bar\lambda_{ij}$=0.028, var($\lambda_{ij}$)=8.60$\times$10$^{-5}$; note that this mean is much lower than that obtained using a prior based on~\citet[Data available from the Dryad Digital Repository: https://doi.org/10.5061/dryad.g7805]{Barbour2016}. Note that, despite the intensive sampling in the~\ref{Kopelke2017} dataset, most \emph{Salix}-galler pairs were never observed co-occurring (3,986/4,992; Fig.~\ref{histograms}).

>>>>

\author{Alyssa R. Cirtwill$^{1,2\dagger}$, \&  Anna Ekl\"{o}f$^{1}$, Tomas Roslin$^{3}$, Kate Wootton$^{3}$, Dominique Gravel$^{4}$}
\date{\small$^1$Department of Physics, Chemistry\\ 
and Biology (IFM)\\ 
Link\"{o}ping University\\
Link\"{o}ping, Sweden\\
\medskip $^2$
Current address: Department of Ecology,\\
Environment, and Plant Sciences (DEEP)\\
Stockholm University\\
Stockholm, Sweden\\
\medskip $^3$ Department of Ecology\\ P.O. Box 7044, Swedish University of Agricultural Sciences \\ SE-750 07 Uppsala, Sweden \\
\medskip$^4$ D\'{e}partement de biologie\\ 
Universit\'{e} de Sherbrooke\\ 
Sherbrooke, Canada\\ 
\medskip
$^\dagger$ Corresponding author:\\
alyssa.cirtwill@gmail.com\\
+46 723 158464\\
}
\renewcommand\Authands{ and }

@@

% \linenumbers

@@

  To demonstrate the effects of nested interaction, process, and detection uncertainty on network structure (see Fig. 1, main text), we can observe how the structure of randomly-generated networks changes as we add layers of uncertainty. We began by constructing 100 ``true" networks with 50 consumer species, 50 resource species, and a connectance of 0.1. We ensured that each network had at least one link per consumer and per resource. For each network, we calculated the mean number of links per consumer, mean number of links per resource, and nestedness (NODF). For simplicity, we assume that there is no interaction uncertainty in this system (i.e., all interactions are either 100\% feasible or 100\% unfeasible based on species' traits).

@@

  With each additional layer of uncertainty, the number of consumers and resources included in the web decreased (mean 49.8 of each species type in detection-filter webs and mean 48.6 of each type in process-filtered webs). The connectance and number of links per consumer and per resource decreased significantly at each level (Fig.~\ref{sim_props}A, B, D; $p$\textless0.001 in all cases based on one-tailed Student's t tests). The effects of uncertainty on nestedness were less predictable, particularly in the detection-filtered webs where variability was much higher than in other web types. Nevertheless, mean nestedness was lower in detection-filtered webs than in the original webs ($t$=19.9, $p$\textless0.001) and lower in the process-filtered webs than the original and detection-filtered webs (Fig.~\ref{sim_props}C; $t$=32.8, $p$\textless0.001 and $t$=93.7, $p$\textless0.001). We note that these tests may reflect the large number of detection-filtered webs generated by our simulation. A single observed network drawn from these detection-filtered webs might well have a higher nestedness than the ``true" network. This simulation demonstrates the fact that process and detection uncertainty mean that we are very unlikely to observe all of the interactions that can truly occur in a system. It also highlights the sometimes unpredictable effects that this uncertainty can have on higher-order structures such as nestedness.

@@

        The obvious rule of thumb to reduce uncertainty is ``sample more". Where resources permit, increasing sample sizes will reduce uncertainty about the upper bound of the interaction probability and will also increase the probability of detecting unlikely or cryptic interactions (e.g., interactions where $L_{ij} = 1$ but process or detection uncertainty is high). Despite these benefits, we note that there are limits to the utility of increased sampling. Since the probability of observing the co-occurrence of two species will always be higher than the probability of observing their interaction (since the probability of interaction is conditional on both interaction partners being present [\citealp{Gravel2018}]), we will accumulate observations of co-occurrences faster than we will accumulate observations of interactions (Fig. 2C, main text). While this improves our understanding of the set of species present in the community, it introduces yet more uncertainty into the matrix of feasible interactions. Note that it is better to identify the set of potential interactions than to miss a species, so sampling effort should not be reduced in order to exclude rare species.

@@

        \vspace{12pt}
        \noindent\emph{
          degree=c(14, 4, 2, 3, 17, 6, 2, 15, 1, 1)}.
          \vspace{12pt}

@@

        \vspace{12pt}
        \noindent\emph{
           int\_probs = c(0.358, 0.110, 0.055, 0.082, 0.468, 0.165, 0.055, 0.412, 0.028, 0.028)}.
          \vspace{12pt}

        The mean of these interaction probabilities is 0.176, approximately two-thirds of the $\lambda_{ij,MLE}$ obtained from the observed data. We can use the distribution of these interaction probabilities as our prior distribution and estimate the uncertainty surrounding our $\lambda_{ij,MLE}$. With some simple R code (function ``calculate\_parameters", \emph{Appendix S7}), we obtain prior parameters $\alpha$=0.998 and $\beta$=4.63. Using these priors in equations~\ref{mean} and~\ref{variance} above (or in the R function ``calculate\_distribution" in \emph{Appendix S7}), we find a prior $\bar\lambda_{ij}$=0.177 and var($\lambda_{ij}$)=0.026. Adding the observed data ($n_{ij}=10$, $k_{ij}=3$) and using the same code, we obtain posterior parameters $\alpha'$=4.00 and $\beta'$=11.6 and a posterior $\bar\lambda_{ij}$=0.256 and var($\lambda_{ij}$)=0.012. Comparing the posterior distribution to the prior, we see that the posterior is closer to the observed data and that the additional data about interactions between species $i$ and $j$ has reduced the variance. We may also wish to calculate a credible interval (analogous to the frequentist confidence interval). This is also quite straightforward in R (see function ``credible\_interval" in \emph{Appendix S7}). In this case, a 95\% credible interval for $\bar\lambda_{ij}$ is (0.080, 0.491).

@@

      The \emph{Salix}-galler-parasitoid meta-network dataset collected 
      by~\citet{Kopelke2017} consists of a single community type sampled across 
      Europe: willow (\emph{Salix}) species, willow-galling sawflies (Hymenoptera, Tenthredinidae, Nematinae, Euurina), and their natural 
      enemies (hymenopteran parasitoids and coleopteran, lepidopteran, dipteran, 
      and hymenopteran inquilines). The data were collected over 29 years 
      (1982-2010) at 374 unique locations across Europe ranging from Sicily to 
      the Arctic. Since the dataset includes repeated visits to a number of sites, the data set 
      contains 641 site visits. Here we take the more conservative approach and pool visits to the same site for a sample size of 374 sub-networks.

      The meta-network consists of 1,173 different interactions 
      between 52 \emph{Salix} nodes, 92 herbivore nodes, and 126 parasitoid 
      nodes. Interactions were determined by dissecting and rearing gall 
      inhabitants from 165,424 galls.

@@

      For species where $n_{ij}=0$ (3,986/4,992 \emph{Salix}-galler pairs in our dataset), estimates for the mean and variance of $\lambda_{ij}$ can be calculated directly from the prior distribution. For the \emph{Salix}-galler network, these parameters were: $\bar\lambda_{ij}$=0.028, var($\lambda_{ij}$)=8.60$\times$10$^{-5}$; note that this mean is much lower than that obtained using a prior based on~\citet[Data available from the Dryad Digital Repository: https://doi.org/10.5061/dryad.g7805]{Barbour2016}. This low mean corresponds to the fact that, despite the intensive sampling in the~\citet{Kopelke2017} dataset, most \emph{Salix}-galler pairs were never observed co-occurring (3,986/4,992; Fig.~\ref{histograms}).
